---
title: "R Notebook"
output: html_notebook
---

# Load the dplyr library for data manipulation
This step loads the `dplyr` package, which will be used for data manipulation tasks such as filtering and mutating columns.
```{r}
library(dplyr)
```

# Read the GVA dataset from CSV
Here we import the quarterly GVA data from a CSV file.
```{r}
data <- read.csv(file = 'data/processed_gva.csv', sep = ';')
```

# Convert columns to numeric
We ensure the `Year` and `GVA` columns are numeric for calculations.
```{r}
data$YEAR <- as.numeric(data$YEAR)
data$GVA <- as.numeric(data$GVA)
```

# Inspect the dataset
Check the data after importing to confirm values.
```{r}
data
```

# Initialize column for GVA with trend removed
Create a new column that will hold the detrended GVA values.
```{r}
data <- dplyr::mutate(data, GVA_no_trend = NA)
```

# View updated dataset
```{r}
data
```

# Remove trend using first differences
We subtract each GVA value from the previous one to remove the trend component.
```{r}
for (i in 1:length(data$GVA_no_trend)) {
  if (i == 1) {
    data$GVA_no_trend[i] <- NA
  } else {
    data$GVA_no_trend[i] <- data$GVA[i] - data$GVA[i - 1]
  }
}
```

# Inspect data after detrending
```{r}
data
```

# Calculate raw quarterly indicators for seasonality
Compute the mean of the detrended values for each quarter to estimate seasonal effects.
```{r}
raw_indicator_q1 <- mean(dplyr::filter(data, QUARTER == 1)$GVA_no_trend, na.rm = TRUE)
raw_indicator_q2 <- mean(dplyr::filter(data, QUARTER == 2)$GVA_no_trend, na.rm = TRUE)
raw_indicator_q3 <- mean(dplyr::filter(data, QUARTER == 3)$GVA_no_trend, na.rm = TRUE)
raw_indicator_q4 <- mean(dplyr::filter(data, QUARTER == 4)$GVA_no_trend, na.rm = TRUE)

raw_indicator_q1
raw_indicator_q2
raw_indicator_q3
raw_indicator_q4
```

# Calculate overall average of raw indicators
```{r}
average_raw_indicator <- mean(c(raw_indicator_q1, raw_indicator_q2, raw_indicator_q3, raw_indicator_q4))

average_raw_indicator
```

# Compute cleaned quarterly indicators
Subtract the overall mean to remove seasonal bias.
```{r}
cleaned_indicator_q1 <- raw_indicator_q1 - average_raw_indicator
cleaned_indicator_q2 <- raw_indicator_q2 - average_raw_indicator
cleaned_indicator_q3 <- raw_indicator_q3 - average_raw_indicator
cleaned_indicator_q4 <- raw_indicator_q4 - average_raw_indicator

cleaned_indicator_q1
cleaned_indicator_q2
cleaned_indicator_q3
cleaned_indicator_q4
```

# Initialize column for GVA with trend and seasonality removed
```{r}
data <- dplyr::mutate(data, GVA_no_trend_seasonality = NA)
```

# Inspect dataset
```{r}
data
```

# Remove seasonality by subtracting cleaned quarterly indicators
```{r}
for (i in 1:length(data$GVA_no_trend_seasonality)) {
  if (data$QUARTER[i] == 1) {
    data$GVA_no_trend_seasonality[i] <- data$GVA_no_trend[i] - cleaned_indicator_q1
  } else if (data$QUARTER[i] == 2) {
    data$GVA_no_trend_seasonality[i] <- data$GVA_no_trend[i] - cleaned_indicator_q2
  } else if (data$QUARTER[i] == 3) {
    data$GVA_no_trend_seasonality[i] <- data$GVA_no_trend[i] - cleaned_indicator_q3
  } else if (data$QUARTER[i] == 4) {
    data$GVA_no_trend_seasonality[i] <- data$GVA_no_trend[i] - cleaned_indicator_q4
  }
}
```

# View data after removing trend and seasonality
```{r}
data
```

# Filter dataset to include only observations from 2017 onwards
```{r}
data <- dplyr::filter(data, data$YEAR >= 2017)
```

# Plot GVA at different stages of data cleaning
Prepare a combined line plot showing raw, detrended, and seasonally adjusted series.
```{r}
y_min <- min(data$GVA, data$GVA_no_trend, data$GVA_no_trend_seasonality, na.rm = TRUE)
y_max <- max(data$GVA, data$GVA_no_trend, data$GVA_no_trend_seasonality, na.rm = TRUE)

plot(1:length(data$GVA), data$GVA, type = "l", col = "blue", lwd = 2, 
     ylim = c(y_min, y_max), xlab = "Time", ylab = "Value", 
     main = "Gross Value Added – Different Stages of Data Cleaning", xaxt = "n")

axis(1, at = 1:length(data$TIME), labels = data$TIME)

par(new = TRUE)
plot(1:length(data$GVA_no_trend), data$GVA_no_trend, type = "l", col = "darkorange", 
     lwd = 2, ylim = c(y_min, y_max), axes = FALSE, xlab = "", ylab = "")

par(new = TRUE)
plot(1:length(data$GVA), data$GVA_no_trend_seasonality, type = "l", col = "darkgreen", 
     lwd = 2, ylim = c(y_min, y_max), axes = FALSE, xlab = "", ylab = "")

legend("topleft", legend = c("GVA (raw)", "GVA (no trend)", "GVA (no trend + seasonality)"), 
       col = c("blue", "darkorange", "darkgreen"), lwd = 2, bty = "n")

```